Ship-based lidar evaluation of Southern Ocean clouds in the storm-resolving general circulation model ICON, and the ERA5 and MERRA-2 reanalyses

### 2025-07-23 manuscript.tex
Rename manuscript

\fontsize{13pt}{15pt}\selectfont

\begin{center} \Large \sffamily\textbf{Ship-based lidar evaluation of Southern Ocean clouds in the storm-resolving general circulation model ICON and the ERA5 and MERRA-2 reanalyses}\\[0.4cm]
\normalsize Peter Kuma$^\mathsf{1,2,\star}$, Frida A.-M. Bender$^\mathsf{1,2}$, Adrian J. McDonald$^\mathsf{3}$, Simon P. Alexander$^\mathsf{4,5}$, Greg M. McFarquhar$^\mathsf{6,7}$, John J. Cassano$^\mathsf{8,9,10}$, Graeme E. Plank$^\mathsf{3}$, Sean Hartery$^\mathsf{11}$, Simon Parsons$^\mathsf{12}$, Sally Garrett$^\mathsf{13}$, and Alex J. Schuddeboom$^\mathsf{3}$\\[0.4cm]
\footnotesize
$^\mathsf{1}$Department of Meteorology (MISU), Stockholm University, Stockholm, Sweden\\
$^\mathsf{2}$Bolin Centre for Climate Research, Stockholm University, Stockholm, Sweden\\
$^\mathsf{3}$School of Physical and Chemical Sciences, University of Canterbury, Christchurch, Aotearoa/New Zealand\\
$^\mathsf{4}$Australian Antarctic Division, Kingston, Tasmania, Australia\\
$^\mathsf{5}$Australian Antarctic Program Partnership, Institute for Marine and Antarctic Studies, University of Tasmania, Hobart, Tasmania, Australia\\
$^\mathsf{6}$Cooperative Institute of Severe and High Impact Weather Research and Operations, University of Oklahoma, Norman, OK, USA\\
$^\mathsf{7}$School of Meteorology, University of Oklahoma, Norman, OK, USA\\
$^\mathsf{8}$Cooperative Institute for Research in Environmental Sciences, University of Colorado, Boulder, CO, USA\\
$^\mathsf{9}$National Snow and Ice Data Center, University of Colorado, Boulder, CO, USA\\
$^\mathsf{10}$Department of Atmospheric and Oceanic Sciences, University of Colorado, Boulder, CO, USA\\[0.2cm]
$^\mathsf{11}$Department of Physics \& Atmospheric Science, Dalhousie University, Halifax, Canada\\
$^\mathsf{12}$New South Wales Department of Planning and Environment, Sydney, New South Wales, Australia\\
$^\mathsf{13}$New Zealand Defence Force, Wellington, New Zealand\\
$^\star$Corresponding author: Peter Kuma (\href{mailto:peter.kuma@misu.su.se}{peter.kuma@misu.su.se})\\[0.4cm]
\normalsize \today\\[0.4cm]

\end{center}

\section*{Abstract}

Global storm-resolving models (GSRMs) are the next avenue of climate modelling.
Among them is a 5-km horizontal resolution version of the Icosahedral Nonhydrostatic Weather and
Climate Model (ICON). This high resolution means that parameterizations of
convection and clouds including subgrid-scale clouds are not required, though microphysics and turbulence
parametrizations are still needed.  Standard-resolution (10--100-km) models, which use
subgrid-scale convection and cloud parametrization, have substantial cloud
biases over the Southern Ocean (SO), adversely affecting radiation and sea surface
temperature.  We evaluated SO clouds in ICON and the ERA5 and MERRA-2
reanalyses.  The SO is dominated by low clouds, which cannot be observed
accurately from space due to overlapping clouds, attenuation, and ground
clutter.  Instead, we analysed about 2400 days of lidar observations from 31
voyages and a station during 2010--2021 using a ground-based lidar simulator.
ICON and the reanalyses underestimate the total cloud fraction by about 10 and
20\%, respectively. ICON and ERA5 overestimate the cloud occurrence peak at
about 500 m, potentially explained by their lifting condensation levels being
too high.  The reanalyses strongly underestimate near-surface clouds or fog.
MERRA-2 tends to underestimate cloud occurrence at all heights.  In daily cloud
cover, ICON and the reanalyses tend to be about 1 and 2 oktas clearer,
respectively. The largest
errors in ICON and the reanalyses occur during periods of weak stability.  Compared to radiosondes, potential
temperature is accurate in the reanalyses to about 0.5 K, but ICON
underestimates stability over the low-latitude SO and is too humid in the boundary
layer. MERRA-2 is too humid at all heights. SO cloud biases remain a
substantial issue in the GSRM, but are an improvement over the lower-resolution
reanalyses. Explicitly resolved convection and cloud processes (except
microphysics) were not enough to address the model cloud biases.

@@

Increasing climate model resolution is one way of improving the accuracy of
the representation of the climate system in models \citep{mauritsen2022}. It has been
practiced since the advent of climate modelling as more computational power,
memory, and storage capacity become available. It is, however, often not as
easy as changing the grid size because of the complex interplay between model
dynamics and physics, which necessitates adjusting and tuning all components
together. Increasing resolution is of course limited by the available
computational power and a trade-off with increasing parameterization
complexity, which is another way of improving model accuracy. Current
computational availability and acceleration from general-purpose computing on
graphics processing units (GPUs) has progressed to enable km-scale (also called
k-scale) Earth system models (ESMs) and coupled atmosphere--ocean general
circulation models (AOGCMs) for research today and will become operational in
the future. Therefore, it represents a natural advance in climate
modelling. Global storm-resolving models (GSRMs) are emerging as a new front
in the development of high-resolution global climate models, with horizontal
grid resolutions of about 2--8 km \citep{satoh2019,stevens2019}. This resolution is enough
to resolve mesoscale convective storms, but smaller-scale convective plumes and
cloud structure remain unresolved. At an approximately 5-km scale,
non-hydrostatic processes also become important \citep{weisman1997}, and for
this reason such models are generally non-hydrostatic. The terms global
cloud-resolving models or global convection-permitting/-resolving models are
also sometimes used interchangeably with GSRMs but imply that clouds or
convection are resolved explicitly, which is not entirely true for GSRMs, as
this would require an even higher horizontal resolution \citep{satoh2019}.
Representative of these efforts is the DYnamics of the Atmospheric general
circulation Modeled On Non-hydrostatic Domains (DYAMOND) project
\citep{stevens2019,dyamond}, which is an intercomparison of nine global GSRMs
over two 40-day time periods in summer (1 August -- 10 September 2016) and
winter (20 January -- 1 March 2020). A new one-year GSRM intercomparison is
currently proposed by \cite{takasuka2024}, with the hope of also evaluating the
seasonal cycle and large-scale circulation.  An alternative to using a
computationally costly GSRM is to train an artificial neural network on GSRM
output and use it for subgrid-scale clouds, as done with the GSRM ICON by
\cite{grundner2022} and \cite{grundner2023}.

The nextGEMS project \citep{nextgems} focuses on the research and development
of GSRMs at multiple modelling centres and universities in Europe. The project
also develops GSRM versions of the Icosahedral Nonhydrostatic Weather and
Climate Model (ICON; \cite{hohenegger2023}), the Integrated Forecasting System
[IFS; \cite{ifs2023}], and their ocean components at eddy-resolving
resolutions: ICON-O \citep{korn2022} coupled with ICON and
Finite-Element/volumE Sea ice-Ocean Model [FESOM; \cite{wang2014}] and Nucleus
for European Modelling of the Ocean [NEMO; \cite{madec2023}] coupled with IFS.
The project has so far produced ICON and IFS simulations with three development
versions called Cycle 1--3 and a pre-final version, with a final production
version planned by the end of the project. nextGEMS is not the only project
developing GSRMs; other GSRMs (or GSRM versions of climate models) currently in
development include: Convection-Permitting Simulations With the E3SM Global
Atmosphere Model [SCREAM; \cite{caldwell2021}], Atmospheric Model [NICAM;
\cite{satoh2008}], Unified Model (UM), eXperimental System for High-resolution
modeling for Earth-to-Local Domain [X-SHiELD; \cite{shield}], Action de
Recherche Petite Echelle Grande Echelle-NonHydrostatic version [ARPEGE-NH;
\cite{bubnova1995,voldoire2017}], Finite-Volume Dynamical Core on the Cubed
Sphere [FV3, \cite{lin2004}], the National Aeronautics and Space Administration
(NASA) Goddard Earth Observing System global atmospheric model version 5
[GEOS5; \cite{putman2011}], Model for Prediction Across Scales [MPAS;
\cite{skamarock2012}], and System for Atmospheric Modeling [SAM;
\cite{khairoutdinov2003}].

Multiple cloud properties have an effect on shortwave (SW) and longwave (LW)
radiation. To first order, the total cloud fraction, cloud phase, and the
liquid and ice water path are the most important cloud properties influencing
SW and LW radiation. These properties are in turn influenced by the
atmospheric thermodynamics, convection and circulation, and both the indirect
and direct effects of aerosols. Second order effects on SW and LW radiation are
associated with the cloud droplet size distribution, ice crystal habit, cloud
lifetime, and direct radiative interaction with aerosols.  In the
6$^\mathrm{th}$ phase of the Coupled Model Intercomparison Project [CMIP6;
\cite{eyring2016}], the cloud feedback has increased relative to CMIP5
\citep{zelinka2020}, which is one of the main reasons for the higher climate
sensitivity of CMIP6 models.

The Southern Ocean (SO) is known to be a problematic region for climate model
biases \citep{schuddeboom2021,hyder2018,cesana2022,zhao2022} due to a lack of
surface and in situ observations and being a lower priority region for
numerical weather prediction (NWP) and climate model development because of its
distance from populated areas. Nevertheless, radiation biases and changes over
an area of its size have a substantial influence on the global climate
\citep{rintoul2011}, such as affecting the Earth radiation balance, ocean heat
and carbon uptake \citep{williams2023}, and the SO is also an important part of the
global ocean conveyor belt \citep{wang2014}. In general, marine clouds have a
disproportionate effect on top of atmosphere (TOA) SW radiation due to the
relatively low albedo of the sea surface.  The relative longitudinal symmetry
of the SO means that model cloud biases tend to be similar across longitudes.

Here, we refer to the SO as ocean regions south of 40°S,
low-latitude SO as 40--55°S and high-latitude SO as south of 55°S. The reason
for this dividing latitude is to split the SO into about two equal zones, as
well as the results by \cite{schuddeboom2021} (Fig. 2b) which show a contrast in
CMIP model radiation biases. \cite{schuddeboom2019} (Fig. 2) and
\cite{kuma2020} (Fig. 3) also show contrasting radiation biases in the Hadley
Centre Global Environmental Model, which is also supported by \cite{cesana2022} which  displays contrasting cloud biases due to the 0°C isotherm reaching the surface at 55°S. The findings
of \cite{niu2024}, however, support a different dividing line of 62°S based on
cloud condensation nuclei concentration.

SO radiation biases have been relatively large and systematic compared to the
rest of the globe since at least CMIP3 \citep{trenberth2010}, and the SO SW
cloud radiative effect (CRE) bias is still positive in eight analysed CMIP6
models analysed by \cite{schuddeboom2021} over the high-latitude SO, whereas
over the low-latitude SO it tends to be more neutral or negative in some
models. Too much absorbed SW radiation over the SO was also identified in the
GSRM SCREAM \cite{caldwell2021}. Compensating biases are possible, such as the
`too few too bright' cloud bias, characterised by too small cloud fraction and
too large cloud albedo \citep{wall2017,kuma2020}, previously described by
\cite{webb2001}, \cite{weare2004}, \cite{zhang2005}, \cite{karlsson2008},
\cite{nam2012}, \cite{klein2013}, and \cite{bender2017} in other regions and
models, which means that a model can maintain a reasonable SW radiation balance
by reflecting too much SW radiation from clouds, but these cover too small
area. A study by \cite{konsta2022} showed that this type of bias is still
present in six analysed CMIP6 models in tropical marine clouds, using the
GCM-Oriented Cloud-Aerosol Lidar and Infrared Pathfinder Satellite Observation
(CALIPSO) Cloud Product [CALIPSO--GOCCP; \cite{chepfer2010}] and Polarization
\& Anisotropy of Reflectances for Atmospheric Sciences coupled with
Observations from a Lidar [PARASOL; \cite{lier2008}] as a reference. They
suggest improper simulation of subgrid-scale cloud heterogeneity as a cause.
Compensating cloud biases in the Australian Community Climate and Earth System
Simulator (ACCESS) – Atmosphere-only model version 2 (AM2) over the SO were
analysed by \cite{fiddes2022} and \cite{fiddes2024}.  \cite{possner2022} showed
that over the SO, the DYAMOND GSRM ICON underestimates low-level cloud fraction
on the order of 30\% [relative to Moderate Resolution Imaging Spectroradiometer
(MODIS) data] and overestimates downwelling TOA SW radiation by approximately
10 Wm$^\mathrm{-2}$ [relative to the Clouds and the Earth’s Radiant Energy
System (CERES) data] in the highest model resolution run (2.5 km).
\cite{zhao2022} reported a similar SW radiation bias in five analysed CMIP6
models over the high-latitude SO and an underestimation of total cloud fraction on
the order of 10\% over the entire 40--60°S SO. Recently, \cite{ramadoss2024}
analysed 48 hours of km-scale ICON limited area model NWP simulations over a SO
region adjacent to Tasmania against the Clouds, Aerosols, Precipitation,
Radiation, and atmospherIc Composition Over the southeRn oceaN (CAPRICORN)
voyage cloud and precipitation observations \cite{mcfarquhar2021}. They found
the ICON cloud optical thickness was underestimated relative to Himawari‐8
satellite observations, but also identified large differences in cloud top
phase.

In general, sea surface temperature (SST) biases in the SO can originate either
in the atmosphere \citep{hyder2018}, caused by too much shortwave heating of
the surface, too little longwave cooling of the surface, or in the ocean
circulation. Interactions of both are also possible, for example SST affecting
clouds and clouds affecting the surface radiation. Using the European Centre
for Medium-Range Weather Forecasts (ECMWF) Reanalysis 5 (ERA5) as a reference,
\cite{zhang2023} have shown that SST biases have improved in CMIP6 compared to
CMIP5, with SST overall increasing in CMIP6. However, over the
SO this resulted in an even higher positive bias, especially in the Atlantic
Ocean (AO) sector of the SO, increasing by up to 1°C.  \cite{luo2023}
identified that the SO SST bias in an ensemble of 18 CMIP6 models originates
not from the surface heat and radiation fluxes (using reanalyses as a
reference), but from a warm bias in the Northern Atlantic Deep Water.

The main aim of this study is to evaluate the GSRM version of ICON. ICON
is developed and maintained jointly by Deutscher Wetterdienst,
Max-Planck-Institute for Meteorology, Deutsches Klimarechenzentrum (DKRZ),
Karlsruhe Institute of Technology, and the Center for Climate Systems Modeling.
Previous studies have identified substantial large-scale biases in climate
model clouds over the SO, affecting sea surface temperature and the Earth’s
albedo. Our aim is to quantify how well the GSRM ICON simulates clouds in
this region, particularly in light of the fact that subgrid-scale clouds and
convection are not parameterized in this model. This region is mostly dominated
by boundary layer clouds generated by shallow convection, and these are
problematic to observe by spaceborne lidars and radars, which are affected by
attenuation by overlapping and thick clouds \citep{mace2009,medeiros2010} and
ground clutter \citep{marchand2008}, respectively.  Specifically, the radar on
CloudSat and lidar on CALIPSO (neither of which are now operational) are
affected by the above mentioned issues, resulting in a strong underestimation of
cloud occurrence below 2 km relative to ground-based lidar observations
\citep{mcerlich2021}.  We hypothesise that this, in turn, can lead to
systematic biases in low clouds in climate models, which are frequently
evaluated against CloudSat--CALIPSO products. Reanalyses can also suffer from
cloud biases, as these are usually parametrised in their atmospheric component,
and also in regions where input observations are sparse. This makes them a
problematic reference for clouds over the SO, and any biases relative to a
reanalysis should be interpreted with caution. Instead, we chose to use a large
set of ship-based observations conducted with ceilometers and lidars on board
the RV \emph{Polarstern} and other voyages and stations as a reference for model evaluation.

Altogether, we analysed about 2400 days of data from 31 voyages, and one
sub-Antarctic station covering diverse longitudes and latitudes of the SO. To
achieve a like-for-like comparison with the model, we used a ground-based lidar
simulator called the Automatic Lidar and Ceilometer Framework [ALCF;
\cite{kuma2021}]. We contrasted the results with ERA5 \citep{era5} and the
Modern-Era Retrospective analysis for Research and Applications, Version 2
[MERRA-2; \cite{gelaro2017}].

\begin{figure}[b!]
\centering
\includegraphics[width=\textwidth]{img/map_fig.pdf}
\caption{
\textbf{(a)} A map showing the tracks of 31 voyages of RV \emph{Polarstern},
RSV \emph{Aurora Australis}, RV \emph{Tangaroa}, RV \emph{Nathaniel B. Palmer},
and HMNZS \emph{Wellington} and one sub-Antarctic station (Macquarie Island)
analysed here. The tracks cover Antarctic sectors south of South America, the
Atlantic Ocean, Africa, Australia, and New Zealand in the years 2010--2021
(inclusive).  The dotted and dashed lines at 40°S and 55°S delineate the
Southern Ocean area of our analysis and its partitioning into two subsets,
respectively.  A photo of \textbf{(b)} RV \emph{Polarstern} (© Folke Mehrtens,
Alfred-Wegener-Institut), \textbf{(c)} Lufft CHM 15k installed on RV
\emph{Tangaroa} (© Peter Kuma, University of Canterbury), \textbf{(d)} Vaisala
CL51 (© Jeff Aquilina, Bureau of Meteorology), \textbf{(e)} Vaisala CT25K at
Macquarie Island (© Simon P.\ Alexander, Australian Antarctic Division).
}
\label{fig:map}
\end{figure}

@@

\subsection{Voyage and station data}

>>>>

\title{Ship-based lidar evaluation of Southern Ocean clouds in the storm-resolving general circulation model ICON, and the ERA5 and MERRA-2 reanalyses}

\authors{Peter Kuma\affil{1,2,3}\thanks{}, Frida A.-M. Bender\affil{1,2}, Adrian J. McDonald\affil{3}, Simon P. Alexander\affil{4,5}, Greg M. McFarquhar\affil{6,7}, John J. Cassano\affil{8,9,10}, Graeme E. Plank\affil{3}, Sean Hartery\affil{3}\thanks{Current affiliation: Department of Physics \& Atmospheric Science, Dalhousie University, Halifax, Canada}, Simon Parsons\affil{3}\thanks{Current affiliation: New South Wales Department of Planning and Environment, Sydney, New South Wales, Australia}, Sally Garrett\affil{11}, and Alex J. Schuddeboom\affil{3}}

\affiliation{1}{Department of Meteorology (MISU), Stockholm University, Stockholm, Sweden}
\affiliation{2}{Bolin Centre for Climate Research, Stockholm University, Stockholm, Sweden}
\affiliation{3}{School of Physical and Chemical Sciences, University of Canterbury, Christchurch, Aotearoa/New Zealand}
\affiliation{4}{Australian Antarctic Division, Kingston, Tasmania, Australia}
\affiliation{5}{Australian Antarctic Program Partnership, Institute for Marine and Antarctic Studies, University of Tasmania, Hobart, Tasmania, Australia}
\affiliation{6}{Cooperative Institute of Severe and High Impact Weather Research and Operations, University of Oklahoma, Norman, OK, USA}
\affiliation{7}{School of Meteorology, University of Oklahoma, Norman, OK, USA}
\affiliation{8}{Cooperative Institute for Research in Environmental Sciences, University of Colorado, Boulder, CO, USA}
\affiliation{9}{National Snow and Ice Data Center, University of Colorado, Boulder, CO, USA}
\affiliation{10}{Department of Atmospheric and Oceanic Sciences, University of Colorado, Boulder, CO, USA}
\affiliation{11}{New Zealand Defence Force, Wellington, Aotearoa/New Zealand}

\correspondingauthor{Peter Kuma}{peter@peterkuma.net}

\begin{keypoints}
\item Ground-based lidar evaluation of a km-scale global climate model and reanalyses reveals substantial cloud biases over the Southern Ocean.
\item Fog or low cloud is underestimated in the reanalyses. In all models, a cloud peak at 500~m tends to be overestimated and too high.
\item A ``too few, too bright'' problem of underestimated cloud fraction, compensated by overestimated cloud albedo, is present in the reanalyses.
\end{keypoints}

\begin{abstract}
Global storm-resolving models (GSRMs) are the latest generation of global climate models. One of them is a 5-km Icosahedral Nonhydrostatic Weather and Climate Model (ICON). Its high resolution means that parameterizations of convection and clouds, including subgrid-scale clouds, are omitted, relying on explicit simulation but still utilizing microphysics and turbulence parameterizations. Standard-resolution (10--100~km) models, which use convection and cloud parameterizations, have substantial cloud biases over the Southern Ocean (SO), adversely affecting radiation and sea surface temperature. The SO is dominated by low clouds, which cannot be observed accurately from space due to overlapping clouds, attenuation, and ground clutter. We evaluated SO clouds in ICON and the ERA5 and MERRA-2 reanalyses using about 2400 days of lidar observations and 2300 radiosonde profiles from 31 voyages and Macquarie Island station during 2010--2021, compared with the models using a ground-based lidar simulator. We found that ICON and the reanalyses underestimate the total cloud fraction by about 10 and 20\%, respectively. ICON and ERA5 overestimate the cloud occurrence peak at about 500 m, potentially explained by their lifting condensation levels being too high. The reanalyses strongly underestimate fog or near-surface clouds, and MERRA-2 underestimates cloud occurrence at almost all heights. Outgoing shortwave radiation is overestimated in the reanalyses, implying a ``too few, too bright'' cloud problem. Thermodynamic conditions are relatively well represented, but ICON is less stable than observations, and MERRA-2 is too humid. SO cloud biases are a substantial issue in the GSRM, but it matches the observations better than the lower-resolution reanalyses.
\end{abstract}

\section*{Plain Language Summary}

Global storm-resolving models are climate models with km-scale horizontal resolution, which are currently in development. Reanalyses are the best estimates of past meteorological conditions based on an underlying global model and observations. We evaluated clouds and thermodynamic profiles over the Southern Ocean in one such model, as well as two reanalyses, based on 2400 days of ship and station observations. Thanks to the high resolution, the model relies entirely on explicit simulation of clouds, instead of subgrid-scale parameterizations. For the evaluation, we used ceilometer and radiosonde observations and a lidar simulator, which enables a fair comparison with the model and reanalyses. We subsetted our results by cyclonic activity and stability. We found that the model and reanalyses underestimate a lidar-derived cloud fraction, and the reanalyses do so more strongly. Fog or near-surface clouds are especially underestimated in the reanalyses. However, the model and one of the reanalyses also tend to overestimate the peak of cloud occurrence at 500 m above the ground, and it tends to be higher. This is linked to thermodynamic profiles, which show a higher lifting condensation level. Southern Ocean biases are still an important problem in the model, but an improvement over the reanalyses is notable.

@@

Increasing climate model resolution is one way of improving the accuracy of the representation of the climate system in models \cite{mauritsen2022}. It has been practiced since the advent of climate modeling as more computational power, memory, and storage capacity become available. It is, however, often not as easy as changing the grid size because of the complex interplay between model dynamics and physics, which necessitates adjusting and tuning all components together. Increasing resolution is, of course, limited by the available computational power and a trade-off with increasing parameterization complexity, which is another way of improving model accuracy. Current computational availability and acceleration from general-purpose computing on graphics processing units (GPUs) has progressed to enable km-scale (also called k-scale) Earth system models (ESMs) and coupled atmosphere--ocean general circulation models (AOGCMs) for research today and will become operational in the future. Therefore, it represents a natural advance in climate modeling. Global storm-resolving models (GSRMs) are emerging as a new front in the development of high-resolution global climate models, with horizontal grid resolutions of about 2--8~km \cite{satoh2019,stevens2019}. This resolution is enough to resolve mesoscale convective storms, but smaller-scale convective plumes and cloud structure remain unresolved. At an approximately 5-km scale, non-hydrostatic processes also become important \cite{weisman1997}, and for this reason such models are generally non-hydrostatic. The terms global cloud-resolving models or global convection-permitting/-resolving models are also sometimes used interchangeably with GSRMs but imply that clouds or convection are resolved explicitly, which is not entirely true for GSRMs, as this would require an even higher horizontal resolution \cite{satoh2019}. Representative of these efforts is the DYnamics of the Atmospheric general circulation Modeled On Non-hydrostatic Domains (DYAMOND) project \cite{stevens2019,dyamond}, which is an intercomparison of nine global GSRMs over two 40-day time periods in summer (1~August--10~September 2016) and winter (20~January--1~March 2020). A new one-year GSRM intercomparison is currently proposed by \citeA{takasuka2024}, with the hope of also evaluating the seasonal cycle and large-scale circulation. An alternative to using a computationally costly GSRM is to train an artificial neural network on GSRM output and use it for subgrid-scale clouds, as done with the GSRM ICON by \citeA{grundner2022} and \citeA{grundner2023}.

The nextGEMS project \cite{nextgems} focuses on the research and development of GSRMs at multiple modeling centers and universities in Europe. The project also develops GSRM versions of the Icosahedral Nonhydrostatic Weather and Climate Model (ICON; \citeA{hohenegger2023}), the Integrated Forecasting System [IFS; \citeA{ifs2023}], and their ocean components at eddy-resolving resolutions: ICON-O \cite{korn2022} coupled with ICON and Finite-Element/volumE Sea ice-Ocean Model [FESOM; \citeA{wang2014a}] and Nucleus for European modeling of the Ocean [NEMO; \citeA{madec2023}] coupled with IFS. The project has so far produced ICON and IFS simulations with three development versions called Cycle 1--3 and a pre-final version, with a final production version planned by the end of the project. nextGEMS is not the only project developing GSRMs; other GSRMs (or GSRM versions of climate models) currently in development include: Convection-Permitting Simulations With the E3SM Global Atmosphere Model [SCREAM; \citeA{caldwell2021}], Atmospheric Model [NICAM; \citeA{satoh2008}], Unified Model (UM), eXperimental System for High-resolution modeling for Earth-to-Local Domain [X-SHiELD; \citeA{shield}], Action de Recherche Petite Echelle Grande Echelle-NonHydrostatic version [ARPEGE-NH; \citeA{bubnova1995,voldoire2017}], Finite-Volume Dynamical Core on the Cubed Sphere [FV3, \citeA{lin2004}], the National Aeronautics and Space Administration (NASA) Goddard Earth Observing System global atmospheric model version 5 [GEOS5; \citeA{putman2011}], Model for Prediction Across Scales [MPAS; \citeA{skamarock2012}], and System for Atmospheric Modeling [SAM; \citeA{khairoutdinov2003}].

Multiple cloud properties have an effect on shortwave (SW) and longwave (LW) radiation. To first order, the total cloud fraction, cloud phase, and the liquid and ice water path are the most important cloud properties influencing SW and LW radiation. These properties are in turn influenced by the atmospheric thermodynamics, convection and circulation, and both the indirect and direct effects of aerosols. Second-order effects on SW and LW radiation are associated with the cloud droplet size distribution, ice crystal habit, cloud lifetime, and direct radiative interaction with aerosols. In the 6$^\mathrm{th}$ phase of the Coupled Model Intercomparison Project [CMIP6; \citeA{eyring2016}], the cloud feedback has increased relative to CMIP5 \cite{zelinka2020}, which is one of the main reasons for the higher climate sensitivity of CMIP6 models.

The Southern Ocean (SO) is known to be a problematic region for climate model biases \cite{schuddeboom2021,hyder2018,cesana2022,zhao2022} due to a lack of surface and in situ observations and being a lower priority region for numerical weather prediction (NWP) and climate model development because of its distance from populated areas. Nevertheless, radiation biases and changes over an area of its size have a substantial influence on the global climate \cite{rintoul2011}, such as affecting the Earth radiation balance, ocean heat, and carbon uptake \cite{williams2023}, and the SO is also an important part of the global ocean conveyor belt \cite{wang2014b}. In general, marine clouds have a disproportionate effect on top-of-atmosphere (TOA) SW radiation due to the relatively low albedo of the sea surface. The approximate longitudinal symmetry of the SO means that model cloud biases tend to be similar across longitudes.

Here, we refer to the SO as ocean regions south of 40°S, low-latitude SO as 40--55°S and high-latitude SO as south of 55°S. The reason for this dividing latitude is to split the SO into about two equal zones, as well as the results by \citeA{schuddeboom2021} (Fig.~2b) which show a contrast in CMIP model radiation biases. \citeA{schuddeboom2019} (Fig.~2) and \citeA{kuma2020} (Fig.~3) also show contrasting radiation biases in the Hadley Centre Global Environmental Model, which is also supported by \citeA{cesana2022} which displays contrasting cloud biases due to the 0°C isotherm reaching the surface at 55°S. The findings of \citeA{niu2024}, however, support a different dividing line of 62°S based on cloud condensation nuclei concentration.

SO radiation biases have been relatively large and systematic compared to the rest of the globe since at least CMIP3 \cite{trenberth2010}, and the SO SW cloud radiative effect (CRE) bias is still positive in eight analyzed CMIP6 models analyzed by \citeA{schuddeboom2021} over the high-latitude SO, whereas over the low-latitude SO it tends to be more neutral or negative in some models. Too much absorbed SW radiation over the SO was also identified in the GSRM SCREAM \cite{caldwell2021}. Compensating biases are possible, such as the ``too few too bright'' cloud bias, characterized by too small cloud fraction and too large cloud albedo \cite{wall2017,kuma2020}, previously described by \citeA{webb2001}, \citeA{weare2004}, \citeA{zhang2005}, \citeA{karlsson2008}, \citeA{nam2012}, \citeA{klein2013}, and \citeA{bender2017} in other regions and models, which means that a model can maintain a reasonable SW radiation balance by reflecting too much SW radiation from clouds, but these cover too small an area. A study by \citeA{konsta2022} showed that this type of bias is still present in six analyzed CMIP6 models in tropical marine clouds, using the GCM-Oriented Cloud-Aerosol Lidar and Infrared Pathfinder Satellite Observation (CALIPSO) Cloud Product [CALIPSO--GOCCP; \citeA{chepfer2010}] and Polarization \& Anisotropy of Reflectances for Atmospheric Sciences coupled with Observations from a Lidar [PARASOL; \citeA{lier2008}] as a reference. They suggest improper simulation of subgrid-scale cloud heterogeneity as a cause. Compensating cloud biases in the Australian Community Climate and Earth System Simulator (ACCESS) – Atmosphere-only model version 2 (AM2) over the SO were analyzed by \citeA{fiddes2022} and \citeA{fiddes2024}. \citeA{possner2022} showed that over the SO, the DYAMOND GSRM ICON underestimates low-level cloud fraction on the order of 30\% and overestimates net downward TOA SW radiation by approximately 10 Wm$^\mathrm{-2}$ in the highest model resolution run (2.5~km). \citeA{zhao2022} reported a similar SW radiation bias in five analyzed CMIP6 models over the high-latitude SO and an underestimation of the total cloud fraction on the order of 10\% over the entire 40--60°S SO. Recently, \citeA{ramadoss2024} analyzed 48~hours of km-scale ICON limited-area model NWP simulations over a SO region adjacent to Tasmania against the Clouds, Aerosols, Precipitation, Radiation, and atmospherIc Composition Over the southeRn oceaN (CAPRICORN) voyage cloud and precipitation observations \cite{mcfarquhar2021}. They found the ICON cloud optical thickness was underestimated relative to Himawari‐8 satellite observations but also identified large differences in cloud top phase.

In general, sea surface temperature (SST) biases in the SO can originate either in the atmosphere \cite{hyder2018}, caused by too much shortwave heating the surface, too little longwave cooling the surface, or in the ocean circulation. Interactions of both are also possible, for example, SST affecting clouds and clouds affecting the surface radiation. Using the European Centre for Medium-Range Weather Forecasts (ECMWF) Reanalysis 5 (ERA5) as a reference, \citeA{zhang2023} have shown that SST biases have improved in CMIP6 compared to CMIP5, with SST overall increasing in CMIP6. However, over the SO this resulted in an even higher positive bias, especially in the Atlantic Ocean (AO) sector of the SO, increasing by up to 1°C. \citeA{luo2023} identified that the SO SST bias in an ensemble of 18 CMIP6 models originates not from the surface heat and radiation fluxes (using reanalyses as a reference), but from a warm bias in the Northern Atlantic Deep Water.

The main aim of this study is to evaluate the GSRM version of ICON. ICON is developed and maintained jointly by Deutscher Wetterdienst, Max-Planck-Institute for Meteorology, Deutsches Klimarechenzentrum (DKRZ), Karlsruhe Institute of Technology, and the Center for Climate Systems Modeling. Previous studies have identified substantial large-scale biases in climate model clouds over the SO, affecting sea surface temperature and the Earth’s albedo. Our aim is to quantify how well the GSRM ICON simulates clouds in this region, particularly in light of the fact that subgrid-scale clouds and convection are not parameterized in this model. This region is mostly dominated by boundary layer clouds generated by shallow convection, and these are problematic to observe by spaceborne lidars and radars, which are affected by attenuation by overlapping and thick clouds \cite{mace2009,medeiros2010} and ground clutter \cite{marchand2008}, respectively. Specifically, the radar on CloudSat and lidar on CALIPSO (neither of which are now operational) are affected by the above-mentioned issues, resulting in a strong underestimation of cloud occurrence below 2 km relative to ground-based lidar observations \cite{mcerlich2021}. We hypothesize that this, in turn, can lead to systematic biases in low clouds in climate models, which are frequently evaluated against CloudSat--CALIPSO products. Reanalyses can also suffer from cloud biases, as these are usually parameterized in their atmospheric component and also in regions where input observations are sparse. This makes them a problematic reference for clouds over the SO, and any biases relative to a reanalysis should be interpreted with caution. Instead, we chose to use a large set of ship-based observations conducted with ceilometers and lidars on board the RV \emph{Polarstern} and other voyages and stations as a reference for the model evaluation.

Altogether, we analyzed about 2400 days of data from 31 voyages and one sub-Antarctic station covering diverse longitudes and latitudes of the SO. To achieve a like-for-like comparison with the model, we used a ground-based lidar simulator called the Automatic Lidar and Ceilometer Framework [ALCF; \citeA{kuma2021}]. We contrasted the results with ERA5 \cite{era5} and the Modern-Era Retrospective analysis for Research and Applications, Version~2 [MERRA-2; \citeA{gelaro2017}].

\begin{figure}[b!]
\centering
\includegraphics[width=\textwidth]{img/map_fig.pdf}
\caption{
\textbf{(a)} A map showing the tracks of 31 voyages of RV \emph{Polarstern}, RSV \emph{Aurora Australis}, RV \emph{Tangaroa}, RV \emph{Nathaniel B. Palmer}, and HMNZS \emph{Wellington} and one sub-Antarctic station (Macquarie Island) analyzed here. The tracks cover Antarctic sectors south of South America, the Atlantic Ocean, Africa, Australia, and New Zealand in the years 2010--2021 (inclusive). The dotted and dashed lines at 40°S and 55°S delineate the Southern Ocean area of our analysis and its partitioning into two subsets, respectively. A photo of \textbf{(b)} RV \emph{Polarstern} (© Folke Mehrtens, Alfred-Wegener-Institut), \textbf{(c)} Lufft CHM~15k installed on RV \emph{Tangaroa} (© Peter Kuma, University of Canterbury), \textbf{(d)} Vaisala CL51 (© Jeff Aquilina, Bureau of Meteorology), \textbf{(e)} Vaisala CT25K at Macquarie Island (© Simon P.\ Alexander, Australian Antarctic Division).
}
\label{fig:map}
\end{figure}

@@

\subsection{Voyage and Station Data}

Together, we analyzed data from 31 voyages of RV \emph{Polarstern}, the resupply vessel (RSV) \emph{Aurora Australis}, RV \emph{Tangaroa}, RV \emph{Nathaniel B. Palmer}, Her (now His) Majesty's New Zealand Ship (HMNZS) \emph{Wellington} and one sub-Antarctic station (Macquarie Island) in the SO south of 40°S between 2010 and 2021. Fig.~\ref{fig:map} shows a map of the campaigns, Table \ref{tab:voyages} lists the campaigns, and Table \ref{tab:voyage-references} lists references where available. Altogether, the analyzed dataset comprised 2421 days of data south of 40°S, but the availability of ceilometer data was slightly lower due to gaps in measurements.
